### 2024-07-16 mathematics.tex
update spinors

>>>>

\subsection{Mathematical Foundations for Spinors}
\paragraph{Quadratic form}
Let~\cite[\S6.3]{Jacobson} $V$ be an $n$-dimensional vector space over a field $K$. A quadratic form $Q$ on $V$ is defined by
\begin{equation}
\begin{split}
   Q\colon V\to K,\qquad
   &\forall k\in K, v\in V, Q(k v)=k^2Q(v),\\
   &B\colon (v,w)\mapsto Q(v+w)-Q(v)-Q(w)\text{~is linear in $v$ and $w$.}
\end{split}
\end{equation}
The bilinear form $B$ satisfies $B(x,x)=2Q(x)$. As long as the characteristic of $F$ is not two, $Q$ is determined by $B$.

Equivalently, $Q$ is characterized by a symmetric ``matrix'' $\hat Q$ such that $Q(v)=v^\TT \hat Q v$.
Focusing on $K=\RR$, $\hat Q$ is a real symmetric matrix and thus diagonalizable. Sylvester's law of inertia states that there exists a basis such that $\hat Q=\diag(1,1,\cdots,1,-1,-1,\cdots,-1)$, where the number of $+1$ and $-1$ are respectively $p$ and $q$.
The pair $(p,q)$ is unique and is called the signature of $Q$, where $\rank \hat Q=p+q$.

Quadratic form induces orthogonality: a basis $\{e_i\}$ of $V$ is called orthogonal [orthonormal] under $Q$ if
\begin{equation}
  B(e_i,e_j)\propto \delta_{ij}\qquad\Bigl[B(e_i,e_j)=c_i\delta_{ij}\text{~with~}c_i\in\{2,0,-2\}\Bigr]
\end{equation}
and $(V,Q)$ has an orthogonal basis as long as the characteristic of $K$ is not two.
If $K$ is a ``spin field'' such as $\RR$ or $\CC$ (or, precisely, $\forall \alpha\in K,\exists \beta\in K$ s.t. $\alpha=\beta^2$ or $-\beta^2$), $(V,Q)$ has an orthonormal basis.

\paragraph{Geometric algebra}
Let~\cite{0907.5356} $V$ be an $n$-dimensional vector space over a field $K$. Let $Q$ be a quadratic form on $V$.
Tensor algebra on $V$ is defined by
\begin{equation}
  T(V) = \bigoplus_{k} V^{\otimes k}
       = K\oplus V\oplus (V\otimes V)\oplus\cdots.
\end{equation}
It has an ideal $K_Q$ generated by $\{v\otimes v - Q(v)\mid v\in V\}$:
\begin{equation}
  K_Q = \left\{\sum_{i} a_i\otimes(v_i\otimes v_i-Q(v_i))\otimes b_i\relmiddle| v_i\in V, a_i,b_i \in T(V)\right\}.
\end{equation}
Then the Geometric algebra $\GeoA(V,Q)$ is defined by $T(V)/K_Q$. This is isomorphic to $\Cl(E, K, q_E)$ defined below.

The product in $\GeoA(V,Q)$ is written by $vw \deq [v\otimes w]=v\otimes w+K_q$. Particularly, $v^2 = Q(v)$ and $vw+wv=B(v,w)$.

\paragraph{Clifford Algebra}
For a finite set $X$, a commutative ring $R$ with unit,
the Clifford algebra $\Cl(X,R,r)$ is defined as the free $R$-module generated by the set $2^X$ of all subset of $X$.
Here, $r\colon X\to R$ is an arbitrary function thought of as a signature.

The Clifford algebra $\mathrm{Cl}$ has
addition $\mathrm{Cl}\times\mathrm{Cl}\to\mathrm{Cl}$,
scalar multiplication $R\times\mathrm{Cl}\to\mathrm{Cl}$, and
product
\begin{equation}
  \mathrm{Cl}\times\mathrm{Cl}\to\mathrm{Cl}\ni AB
  \deq \tau(A,B)A\setdiff B,
\end{equation}
where $A\setdiff B=(A\cup B)\setminus(B\cap A)$,
defined based on the following map $\tau\colon2^X\times2^X\to R$:
\begin{equation}
  \begin{split}
  &\tau(\{x\},\{x\}) = r(x)\quad\forall x\in X,&
  &\tau(A,B)\tau(A\setdiff B,C)=\tau(A,B\setdiff C)\tau(B,C)\quad\forall A,B,C\in 2^X,\\
  &\tau(\{y\},\{x\}) = -\tau(\{x\},\{y\})\quad\forall x,y\in X,x\neq y,&
  &\tau(\emptyset,A)=\tau(A,\emptyset)=1\quad\forall A\in2^X,\\
  &\tau(A,B)\in\{-1,1\}\quad \text{if}~A\cap B=\emptyset.
  \end{split}
\end{equation}

Several standard operations are defined: for $A\in2^X$ and $k=\binom{|A|}{2}$,
\begin{equation}
\text{Grade involution}\quad A^\star\deq(-1)^{|A|}A,\qquad
\text{Reversion}\quad A^\dagger\deq(-1)^{k}A,\qquad
\text{Clifford conjugate}\quad A^\mdlgwhtsquare\deq A^{\star\dagger}.
\end{equation}
The sign of grade involution (reversion) is negative iff $|A|$ mod 4 is 1 or 3 (2 or 3).

Let $(V,Q)$ be an $n$-dimensional vector space $V$ over a field $K$ equipped with a quadratic form $Q$. Let $E=\{e_i\}$ be an orthogonal basis of $(V,Q)$.
Then
\begin{equation}
  \Cl(E, K, Q|_E) \cong \GeoA(V,Q).
\end{equation}
The Clifford algebra $\Cl(X,R,r)$ has subspaces of $k$-vectors $\mathrm{Cl}^k$ and of even and odd vectors $\mathrm{Cl}^\pm$:
\[
\Cl(X,R,r)=\mathrm{Cl}^+\oplus\mathrm{Cl}^-=
\mathrm{Cl}^0\oplus\mathrm{Cl}^1\oplus\mathrm{Cl}^2\oplus\cdots\oplus\mathrm{Cl}^{|X|}.
\]
In particular, $\mathrm{Cl}^+$ is a subalgebra.

\paragraph{Groups in Geometric Algebra}
The following groups are embedded in $\GeoA(V,Q)$,
where $V^\times=\{v\in V\mid Q(v)\neq 0\}$:
\begin{align}
  &\GeoA^\times \deq \{g\in\GeoA\mid \exists g^{-1}\in\GeoA, gg^{-1}=g^{-1}g=1\},
  \qquad \Gamma\deq\{v_1v_2\cdots v_n\mid v_i\in V^\times\},
  \qquad \tilde \Gamma\deq \{g\in\GeoA^\times\mid gVg^{-1}\in V\},
\notag\\
  &\Pin\deq\{x\in \Gamma\mid xx^\dagger = \pm1\},
 \qquad\Spin\deq \Pin\cap\GeoA^+,
 \qquad\Spin^+\deq \{x\in\Spin\mid xx^\dagger=1\},
\end{align}
The Lipschitz group $\tilde\Gamma$ is actually equal to the versor group $\Gamma$ and is the smallest group that contains $V^\times$. To summarize,
\[\GeoA\supset\GeoA^\times\supset \Gamma=\tilde\Gamma\supset V^\times;\qquad\Gamma\supset\Pin\supset\Spin\supset\Spin^+.\]

There are surjective homomorphisms with kernel $\pm1$ for
\begin{equation}
  \Pin(s,t)\to\gO(s,t),\qquad
  \Spin(s,t)\to\gSO(s,t),\qquad
  \Spin^+(s,t)\to\gSO^+(s,t).
\end{equation}
Also, for our interest, there are isomorphisms~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor}
\begin{equation}
  \GeoA(\RR^{4,1})\cong \GeoA(\RR^{1,3})\otimes \CC \cong \GeoA(\CC^4)\cong\CC^{4\times4},\qquad
  \Spin^+(1,3)\cong\gSL(2,\CC)\cong\gSp(2,\CC),\qquad
  L_0\cong\mathop{\mathrm{PSL}}(2;\CC)=\gSL(2;\CC)/\ZZ_2.
\end{equation}
Meanwhile, the Lorentz algebra $\aSO(1,3)$ is isomorphic to $\aSL(2;\mathbb C)$ viewed as a real Lie albegra~\cite[\S7.8]{Hall2015},
and its complexification
$\aSO(1,3)_{\mathbb C}$ is isomorphic to $
\aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}
=\aSL(2,\mathbb C)\oplus\aSL(2,\mathbb C)$.
